---
title: "Final Project"
authors: "Johnson He, Anson Yamvinij, Al Gillani, Ani Kondor "
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
editor: source
execute: 
  error: true
  echo: false
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse)
library(glue)
library(english)
library(broom)
library(gridExtra)
library(kableExtra)
install.packages("gganimate")
library(gganimate)
install.packages("plotly")
library(plotly)
library(shiny)
```

## About the Data

Life Expectancy Data Description: The number of years a newborn infant would live if the current mortality rates at different ages were to stay the same throughout its life. The main key is country and the life expectancy is under columns by year.

Happiness Data Description: This is the national average response to the question of life evaluations asking the following "Please imagine a ladder, with steps numbered from 0 at the bottom to 10 at the top. The top of the ladder represents the best possible life for you and the bottom of the ladder represents the worst possible life for you. On which step of the ladder would you say you personally feel you stand at this time?" This measure is also referred to as Cantril life ladder. Gapminder has converted this indicator's scale from 0 to 100 to easly communicate it in terms of percentage. The main key is country and happiness score are under columns by year.

## Hypothesis

We hypothesize that Happiness score will have a direct positive correlation with Life Expectancy, i.e. For each year, the higher the life expectancy, the higher happiness score. Of course, there will be some outliers, but for the most part we think that our hypothesis will be true.

```{r}
hapi <- read.csv("hapiscore_whr.csv")
life <- read.csv("lex.csv")
```

```{r}
colnames(hapi)[1] = "country"
colnames(life)[1] = "country"
hapi<-hapi |>
  mutate(across(.cols=X2005:X2022, .fns = as.numeric)) 
life<-life |>
  select(country,X2005:X2022) |>
  mutate(across(.cols=X2005:X2022, .fns = as.numeric)) 
```

## Data Cleaning Process

All of the happiness scores and life expectancies are numbers, so we converted these values to the numeric type. Additionally, the happiness score data set only contains data from the years 2005-2022. So, we only selected those rows from the life expectancy data set so that they will have an equal amount of columns and years.

```{r}
hapiscore <- hapi %>% 
                pivot_longer(cols = -country, 
                             names_to = "year", 
                             values_to = "happiness_score")

lex <- life %>% 
          pivot_longer(cols = -country, 
                       names_to = "year", 
                       values_to = "lex")

```

```{r}
data <- hapiscore %>%
    inner_join(lex, by = c("country","year"))
```

## Data Visualization

```{r}
#2.1
data |>
  ggplot(aes(x=lex, y= happiness_score)) +
  geom_point() +
  labs(x = "Life Expectancy", y = '', subtitle = "Happiness Score", title = "Life Expectancy v.s Happiness Score (All Years)")
```

```{r}
p <- data %>%
  ggplot(aes(x =lex, y = happiness_score)) +
  geom_jitter() + 
  geom_smooth(method = "lm") +
  labs(x = "Life Expectancy", y="", subtitle="Happiness Score", title="Average Life Expectancy v.s Average Happiness Score")

# Change to interactive
ggplotly(p)

```

```{r}
#2.1
data |>
  mutate( year = str_remove(year, 'X')) |>
  group_by(year)|>
  ggplot(aes(x=lex, y= happiness_score)) +
  geom_point() +
  facet_wrap(vars(year)) +
  labs(x = "Life Expectancy", y = '', subtitle = "Happiness Score", title = "Life Expectancy v.s Happiness Score (Sep. by Years)")

```

```{r}
ui <- fluidPage(
  titlePanel("Interactive Plot: Life Expectancy vs Happiness Score"),
  sidebarLayout(
    sidebarPanel(
      selectInput(inputId = "year",
                  label = "Choose a year:",
                  choices = unique(as.character(data$year)),
                  selected = "X2022")
    ),
    mainPanel(
      plotlyOutput(outputId = "scatterplot")
    )
  )
)
server <- function(input, output) {
  output$scatterplot <- renderPlotly({
    data_specific_year <- data %>% filter(year == input$year)
    p <- data_specific_year %>%
      plot_ly(x = ~lex, y = ~happiness_score, type = 'scatter', mode = 'markers') %>%
      layout(title = paste("Interactive Plot: Life Expectancy vs Happiness Score for", input$year),
             xaxis = list(title = "Life Expectancy"),
             yaxis = list(title = "Happiness Score"))
    p
  })
}

shinyApp(ui = ui, server = server)
```

```{r}


```

From the individual Life Expectancy vs Happiness Score plots we see a positive, linear trend with no significant outliers. However we noticed that there is not much data from 2005. However, from 2006-2022, we see a trend that as life expectancy increases, the happiness score also increases. This is similar to the result we found in the overall Life Expectancy vs Happiness Score plot, where life expectancy and happiness score have a positive correlation.

```{r}
#2.2
#lex is life expectancy
newdata <- data |>
  group_by(country) |>
  summarize(hscore = mean(happiness_score, na.rm=TRUE), lex = mean(lex))
newdata |>
  ggplot(aes(x =lex, y = hscore)) +
  geom_jitter() + 
  geom_smooth(method = "lm") +
  labs(x = "Life Expectancy", y="", subtitle="Happiness Score", title="Average Life Expectancy v.s Average Happiness Score")
SLRM <- lm(hscore ~ lex,data=newdata)
vartab <- SLRM |>
  augment() |>
  summarize(residResponse = format(round(var(.resid),2), nsmall = 2),
            residFitted = format(round(var(.fitted),2), nsmall = 2),
            residReg = format(round(var(hscore),2), nsmall = 2))
kable(vartab, align = "ccc")|>
  kable_styling(latex_options = "striped") |>
  row_spec(1, bold = TRUE, color = 'steelblue') |>
  add_header_above(c('Variance' = 3))
#  summarize(residResponse = var(resid(SLRM)),
#           residFitted = var(fitted.values(SLRM)),
#            residReg = var(newdata$hscore))
#use plot to further analyze SLRM
plot(SLRM)
```

In our plots, we use Life Expectancy score as the explanatory variable and Happiness Score as the response variable. As in **the life expectancy score is x** and **the happiness score is y**. Using a simple linear regression model we get the equation: $$ y = -20.1 + 1.03x$$ The intercept term (-20.1) tells us the predicted happiness score for someone with a life expectancy score of zero. The coefficient for x (1.03) tells us that for every one point increase in life expectancy score, there is a predicted increase of 1.03 in happiness score.

The variance in the fitted values of our regression model are 69.17. Given the variance of the residual of our response variable, 43.47, we are not concerned with our model making inaccurate predictions because the variances do not differ by much and looking at the graph we see variance matches the "band" where most values fall into. The Residual vs Fitted plot does cause some concern with unequal variance, however, given the sample size we do not believe this will be a significant issue. The Normal Q-Q plot also shows an approximately normal distribution for the residuals with a slight curve at the extremities meaning there are a few more extreme values than would be expected if the residuals were truly from a normal distribution. The red band in Scale-Location plot is horizontal indicating homoscedasticity and there is no pattern in the residuals, indicating equal variability at all fitted values. In the Residuals vs Leverage plot there are no observations that fall outside of Cook's distance and therefore are not significantly influential in the regression. The r-squared of our model is 0.6141 meaning 61.41% of the variation in our model is explained by the life expectancy variable. For having only one parameter we feel this is a sufficient score.

```{r}
# 3.1 Visualizing Simulations from the Model
# Remove duplicate columns
newdata <- newdata[!duplicated(names(newdata))]


set.seed(123) # for reproducibility

# Generate predictions
newdata$predicted_hscore <- predict(SLRM, newdata)

# Add random errors to the predictions
newdata$resid <- residuals(SLRM)
newdata$simulated_hscore <- newdata$predicted_hscore + rnorm(nrow(newdata), mean=0, sd=sigma(SLRM))

# Plot observed data
p1 <- ggplot(newdata, aes(x=lex, y=hscore)) +
  geom_point() +
  labs(x = "Life Expectancy", y="",subtitle = 'Happiness Score', title = "Observed Data")

# Plot simulated data
p2 <- ggplot(newdata, aes(x=lex, y=simulated_hscore)) +
  geom_point() +
  labs(x = "Life Expectancy", y = "", subtitle= 'Simulated Happiness Score', title = "Simulated Data")

# Arrange the plots side by side
grid.arrange(p1, p2, ncol=2)
```

```{r}
#str(newdata)
#names(newdata)
#table(duplicated(names(newdata)))
```

```{r}
# Observed Data
p1 <- ggplot(newdata, aes(x=lex, y=hscore)) +
  geom_point() +
  labs(x = "Life Expectancy", y="",subtitle = 'Happiness Score', title = "Observed Data")

# Simulated Data
p2 <- ggplot(newdata, aes(x=lex, y=simulated_hscore)) +
  geom_point() +
  labs(x = "Life Expectancy", y = "", subtitle= 'Simulated Happiness Score', title = "Simulated Data")

# Change to interactive
ggplotly(p1)
ggplotly(p2)

```

```{r}
set.seed(123) # for reproducibility

# Initialize a vector to store R2 values
r2_values <- vector()

# Generate 1000 simulated datasets
for(i in 1:1000){
  # Simulate happiness scores
  simulated_hscore <- newdata$predicted_hscore + rnorm(nrow(newdata), mean=0, sd=sigma(SLRM))
  
  # Fit a linear regression model
  model <- lm(newdata$hscore ~ simulated_hscore)
  
  # Store the R2 value
  r2_values[i] <- summary(model)$r.squared
}

# Plot the distribution of the R2 values
hist(r2_values, main="Distribution of R2 values", xlab="R2 values")
```

```{r}
# Create a data frame with the R2 values
df_r2 <- data.frame(r2_values = r2_values)

# Create the histogram with ggplot2
p <- ggplot(df_r2, aes(x = r2_values)) +
  geom_histogram(binwidth = 0.01, fill = 'lightblue', color = 'black') +
  labs(x = "R2 values", title = "Distribution of R2 values")

# Convert to a plotly object
ggplotly(p)


```
